Imputation of Incomplete Multilevel Data with R

### 2024-01-10 Manuscript/manuscript.tex
Updated manuscript
I have removed the 'no pmm' part from the Heckman model. Additionally, I've included the plot function with echo=FALSE, making it suitable for copying and pasting into the appendix later on. However, the pattern function with cluster still does not work. Regarding the box 3 about the prediction matrix, previously i have added relevant information about this (with an equation) in the Imputation flow. Please review this information to ensure that the box is constructed appropriately, avoiding redundant information throughout the paper. Thanks.

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, breakable, sharp corners, interior hidden]}{\end{tcolorbox}}\fi

@@

The popular \pkg{mice} package in \proglang{R} \citet{R}

\hypertarget{scope}{%
\subsection{scope}\label{scope}}

@@

\hypertarget{concepts-in-multilevel-data}{%
\subsection{concepts in multilevel
data}\label{concepts-in-multilevel-data}}

Many datasets include individuals that are clustered together, for
example in geographic regions, or even different studies. In the
simplest case, individuals (e.g., students) are nested within a single
cluster (e.g., school classes). More complex clustered structures may
occur when there are multiple hierarchical levels (e.g., students in
different schools or patients within hospitals within regions across
countries), or when the clustering is non-nested (e.g., electronic
health record data from diverse settings and populations within large
databases). With clustered data we generally assume that individuals
from the same cluster tend to be more similar than individuals from
other clusters. In statistical terms, this implies that observations
from the same cluster are not independent and may in fact be correlated.
If this correlation is left unaddressed, estimates of \emph{p} values,
confidence intervals even model parameters are prone to bias
\citep{loca01}. Statistical methods for clustered data typically adopt
hierarchical models that explicitly describe the grouping of
observations. These models are also known as `multilevel models',
`hierarchical models', `mixed effect models', `random effect models',
and in the context of time-to-event data as `frailty models'. Table
\ref{tab:clus} provides an overview of some key concepts in multilevel
modeling.

\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, colframe=quarto-callout-color-frame, opacityback=0, colback=white, rightrule=.15mm, left=2mm]

Box 1. The intraclass correlation coefficient.

In \proglang{R}, multilevel models may be fitted using the package
\pkg{lme4}. For linear mixed-effects models, the function

@@

\hypertarget{concepts-in-missing-data}{%
\subsection{concepts in missing data}\label{concepts-in-missing-data}}

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-3-1.pdf}

@@

\hypertarget{imputation-with-mice}{%
\subsection{imputation with mice}\label{imputation-with-mice}}

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, colframe=quarto-callout-color-frame, opacityback=0, colback=white, rightrule=.15mm, left=2mm]

@@

\begin{tcolorbox}[enhanced jigsaw, breakable, toprule=.15mm, arc=.35mm, bottomrule=.15mm, leftrule=.75mm, colframe=quarto-callout-color-frame, opacityback=0, colback=white, rightrule=.15mm, left=2mm]

@@

Before diving into imputation, make sure your estimation model can
handle missing data. Sometimes, simpler methods like complete case
analysis might be suffice, especially if your missingness is low
(usually \textless5\%). There are scenarios in which specific Maximum
Likelihood (ML) estimation methods outperform Multiple Imputation (MI)
methods, for instance when the response variable is the sole incomplete
variable, mixed models demonstrate robustness to missing data under the
Missing at Random (MAR) assumption and with a correct
variance-covariance specification.\cite{Molenberghs_2007}.

@@

The imputation model can incorporate additional terms compared to the
main model without causing compatibility issues. For instance, it's
recommended to include the outcome variable in the imputation model for
prediction variables \cite{moons2006a}. In cases where the outcome is
time-to-event, the Nelson-Aalen estimate of the time to the event should
be added as a covariate in the imputation model {[}REF{]}. Additionally,
including auxiliary variables, even if not part of the main model, can
be linked to the probability of missingness, improving the likelihood of
meeting the Missing at Random (MAR) assumption and enhancing estimation
efficiency \cite{hardt2012a}.

@@

Recipes have been proposed for imputing incomplete level-1 and level-2
variables for hierarchical models \cite{buuren2018a} (\(\S\) 7.10),
which can be convenient to follow when dealing with numerous variables
(interaction terms at different levels) and models with many random
effects that are prone to convergency problems due to overspecification
in the imputation model. These rules were designed to ensure
compatibility among the conditionally specified imputation models and
congeniality between the imputation and the main model. However, they
may not be applicable to all 2l imputation methods; for instance, the
2l.2stage imputation methods of the micemd package only allow the
inclusion of random predictor variables (2).

On the other hand, the formula option is useful in specifying complex
imputation models with polynomial terms or interactions and compared
with the prediction matrix method do not requires the inclusion of
additional terms as Just Another Variable (JAV)
\cite{buuren2018a}(\(\S\) 6.4).

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-8-1.pdf}

@@

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-11-1.pdf}

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-17-1.pdf}

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-22-1.pdf}

@@

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-29-1.pdf}

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-38-1.pdf}

@@

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-41-1.pdf}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1605}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1111}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1235}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0988}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1358}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2222}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0741}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0741}}@{}}
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Std.Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
t.value
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
df
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
P(\textgreater\textbar t\textbar)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
RIV
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
FMI
\end{minipage} \\
\midrule()
\endhead
(Intercept) & -1.892 & 0.333 & -5.682 & 11445.455 & 0.000 & 0.019 &
0.019 \\
type & -0.346 & 0.178 & -1.948 & 686356.768 & 0.051 & 0.002 & 0.002 \\
age & 0.032 & 0.002 & 19.564 & 2418.910 & 0.000 & 0.042 & 0.041 \\
motor\_score2 & -0.575 & 0.070 & -8.226 & 71295.018 & 0.000 & 0.008 &
0.008 \\
motor\_score3 & -0.891 & 0.072 & -12.440 & 66712.189 & 0.000 & 0.008 &
0.008 \\
motor\_score4 & -1.276 & 0.074 & -17.305 & 18426.087 & 0.000 & 0.015 &
0.015 \\
pupil2 & 1.315 & 0.068 & 19.339 & 110.826 & 0.000 & 0.235 & 0.204 \\
pupil3 & 0.657 & 0.075 & 8.745 & 423.463 & 0.000 & 0.108 & 0.101 \\
ct2 & 0.747 & 0.084 & 8.884 & 41.781 & 0.000 & 0.448 & 0.340 \\
ct3 & 0.766 & 0.090 & 8.535 & 12.103 & 0.000 & 1.352 & 0.631 \\
\bottomrule()
\end{longtable}

@@

\begin{longtable}[]{@{}lr@{}}
\toprule()
& Estimate \\
\midrule()
\endhead
Intercept\textasciitilde\textasciitilde Intercept\textbar name &
0.082 \\
\bottomrule()
\end{longtable}

\hypertarget{obesity-data}{%
\subsection{obesity data}\label{obesity-data}}

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-48-1.pdf}

@@

\begin{verbatim}
R> Nulmodel <- lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
R> # performance::icc(Nulmodel)
\end{verbatim}

@@

It suggests using `2l.2stage.bin' for the FAV variable and
`2l.2stage.norm' for the age variable. However, after inspecting the age
density plot, we consider modifying its method to `2l.2stage.pmm'. For
the BMI variable, we employ deterministic imputation.

\begin{verbatim}
meth_mar <-
  find.defaultMethod(
    Obesity,
    ind.clust = 1,
    I.small = 7,
    ni.small = 100,
    prop.small = 0.4
  )
meth_mar["BMI"] <- "~ I(Weight / (Height)^2)"
meth_mar["Age"] <- "2l.2stage.pmm"
\end{verbatim}

@@

\begin{verbatim}
pred_mar <- quickpred(Obesity)
pred_mar[, "Cluster"] <- -2 
pred_mar[, "Time"] <- 0
pred_mar[pred_mar == 1] <- 2
pred_mar[c("Height", "Weight"), "BMI"] <- 0
plot_pred(pred_mar)
\end{verbatim}

{\centering \includegraphics[width=0.7\textwidth,height=\textheight]{manuscript_files/figure-pdf/obesity-predmar-1.pdf}

\begin{verbatim}
imp_mar <-
  mice(
    data = Obesity,
    method = meth_mar,
    predictorMatrix = pred_mar,
    printFlag = FALSE
  )
\end{verbatim}

\begin{verbatim}
summary(complete(imp_mar,"long")$Weight)
\end{verbatim}

\begin{verbatim}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  4.428  69.890  82.800  82.835  95.491 145.885 
\end{verbatim}

We are also contemplating the utilisation of the predictive mean
matching (pmm) option, as the values imputed using a fully parametric
method may be implausibly low for some patients.

\begin{verbatim}
meth_mar["Weight"] <- "2l.2stage.pmm"
imp_mar_pmm <- mice(
  data = Obesity,
  method = meth_mar,
  predictorMatrix = pred_mar,
  printFlag = FALSE
)
\end{verbatim}

\begin{verbatim}
summary(complete(imp_mar_pmm, "long")$Weight)
\end{verbatim}

\begin{verbatim}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  28.35   71.04   83.65   83.87   95.66  134.61 
\end{verbatim}

\begin{verbatim}
plot_trace(imp_mar_pmm, "Weight")
\end{verbatim}

@@

\begin{verbatim}
ggmice(imp_mar_pmm, aes(.imp, Weight)) + 
  geom_jitter()
\end{verbatim}

@@

\begin{verbatim}
R> # fit <- with(
R> #     imp_mar_pmm,
R> #     lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
R> #     )
R> fit <- with(
+    imp_mar_pmm,
+    lmer(BMI ~ Age + FamOb + Gender + (1 + Age | Cluster))
+    )
R> est <- testEstimates(as.mitml.result(fit), extra.pars = TRUE)
\end{verbatim}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1558}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1169}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1299}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1039}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1039}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2338}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0779}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0779}}@{}}
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Std.Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
t.value
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
df
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
P(\textgreater\textbar t\textbar)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
RIV
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
FMI
\end{minipage} \\
\midrule()
\endhead
(Intercept) & 28.445 & 2.259 & 12.592 & 157.553 & 0.000 & 0.190 &
0.170 \\
Age & 0.041 & 0.032 & 1.311 & 388.953 & 0.191 & 0.113 & 0.106 \\
FamObyes & 0.000 & 0.316 & -0.001 & 207.648 & 0.999 & 0.161 & 0.147 \\
GenderMale & -1.361 & 0.411 & -3.310 & 7.986 & 0.011 & 2.421 & 0.761 \\
\bottomrule()
\end{longtable}

\begin{longtable}[]{@{}lr@{}}
\toprule()
& Estimate \\
\midrule()
\endhead
Intercept\textasciitilde\textasciitilde Intercept\textbar Cluster &
19.805 \\
Age\textasciitilde\textasciitilde Age\textbar Cluster & 0.003 \\
Intercept\textasciitilde\textasciitilde Age\textbar Cluster & -0.069 \\
Residual\textasciitilde\textasciitilde Residual & 26.010 \\
ICC\textbar Cluster & 0.394 \\
\bottomrule()
\end{longtable}

@@

Then we proceed to run the imputation model as before, after executing
these imputation procedures, it is essential to assess convergence and
the coherence of the imputed values.

\begin{verbatim}
imp_mnar <- mice(
  data = Obesity,
  method = meth_mnar,
  predictorMatrix = pred_mnar,
  printFlag = FALSE
)
summary(complete(imp_mnar, "long")$Weight)
\end{verbatim}

\begin{verbatim}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  16.31   72.14   86.97   87.68  102.43  190.44 
\end{verbatim}

Upon examining the weight variable, we noticed that the imputed range
falls outside the realm of plausible values (as weight should be
positive).

@@

\begin{verbatim}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  16.31   72.14   86.97   87.68  102.43  190.44 
\end{verbatim}

Consequently, as before we use the `pmm', option but this time for the
Heckman imputation, this approach ensures that the imputed values remain
within the range of observable values. We then run the imputation model
but this time using the option of pmm, to assure that weight values are
in the range of the observable data, this can be implemented by setting
the pmm parameter to true.

\begin{verbatim}
imp_mnar_pmm <-
  mice(
    data = Obesity,
    method = meth_mnar,
    predictorMatrix = pred_mnar,
    pmm = T,
    printFlag = FALSE
  )
\end{verbatim}

\begin{verbatim}
summary(complete(imp_mnar_pmm, "long")$Weight)
\end{verbatim}

\begin{verbatim}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  28.35   71.87   85.58   86.51  100.15  134.61 
\end{verbatim}

\begin{verbatim}
plot_trace(imp_mar_pmm, "Weight")
\end{verbatim}

@@

After this modification we proceed to compare the effects on the model.
We run the analysis model on each of the completed datasets as well as
the dataset where the incomplete values are removed (Complete Case
analysis, CC).

\begin{verbatim}
# cc_rs <- with(setDT(Obesity)[complete.cases(Obesity), ],
#               lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age |
#                     Cluster))
mar_rs <- with(
  imp_mar, 
  lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
  )
mar_pmm_rs <- with(
    imp_mar_pmm,
    lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
    )
mnar_rs <- with(
  imp_mnar, 
  lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
  )
mnar_pmm_rs <- with(
  imp_mnar_pmm,
  lme(BMI ~ Age + FamOb + Gender, random =  ~ 1 + Age | Cluster)
  )
# list_models <- list(cc_rs, mar_rs, mar_pmm_rs, mnar_rs, mnar_pmm_rs)
# plot_models(list_models,
#             mod_name = c("Complete case", "MAR", "MAR_pmm", "MNAR", "MNAR_pmm"))
\end{verbatim}

\begin{verbatim}
est <- testEstimates(as.mitml.result(mnar_pmm_rs), extra.pars = TRUE)
\end{verbatim}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1579}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1184}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1316}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1053}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0921}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2368}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0789}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0789}}@{}}
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Std.Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
t.value
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
df
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
P(\textgreater\textbar t\textbar)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
RIV
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
FMI
\end{minipage} \\
\midrule()
\endhead
(Intercept) & 29.199 & 2.766 & 10.555 & 55.483 & 0.000 & 0.367 &
0.294 \\
Age & 0.050 & 0.038 & 1.301 & 18.731 & 0.209 & 0.859 & 0.512 \\
FamObyes & 0.016 & 0.445 & 0.036 & 15.910 & 0.972 & 1.006 & 0.554 \\
GenderMale & -1.408 & 0.696 & -2.024 & 5.128 & 0.097 & 7.563 & 0.912 \\
\bottomrule()
\end{longtable}

>>>>

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, frame hidden, boxrule=0pt, breakable, enhanced]}{\end{tcolorbox}}\fi

@@

The popular \pkg{mice} package in \proglang{R} \citet{R} TODO:
\citet{Hanne}(Is this section relevant?)

\hypertarget{scope}{%
\subsection{Scope}\label{scope}}

@@

\hypertarget{concepts-in-multilevel-data}{%
\subsection{Concepts in multilevel
data}\label{concepts-in-multilevel-data}}

Many datasets include individuals that are clustered together, for
example in geographic regions, or even different studies. In the
simplest case, individuals (e.g., students) are nested within a single
cluster (e.g., school classes). More complex clustered structures may
occur when there are multiple hierarchical levels (e.g., students in
different schools or patients within hospitals within regions across
countries), or when the clustering is non-nested (e.g., electronic
health record data from diverse settings and populations within large
databases). With clustered data we generally assume that individuals
from the same cluster tend to be more similar than individuals from
other clusters. In statistical terms, this implies that observations
from the same cluster are not independent and may in fact be correlated.
If this correlation is left unaddressed, estimates of \emph{p} values,
confidence intervals even model parameters are prone to bias
\citep{loca01}. Statistical methods for clustered data typically adopt
hierarchical models that explicitly describe the grouping of
observations. These models are also known as `multilevel models',
`hierarchical models', `mixed effect models', `random effect models',
and in the context of time-to-event data as `frailty models'. Table
\ref{tab:clus0} provides an overview of some key concepts in multilevel
modeling.

\begin{table}

\caption{Concepts in multilevel methods}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{3cm}>{\raggedright\arraybackslash}p{12cm}}
\toprule
Concept & Details\\
\midrule
Sample units & Units of the population from which measurements are taken in a sample, e.g., students.\\
Cluster & Variable that specify the cluster or agruppation, e.g., Classroom\\
Hierarchical data & Data are grouped into clusters at different levels, observations belonging to the same cluster are expected to share certain characteristics.\\
Level-1 & Variable that varies within a cluster, eg. Test score\\
Level-2 & Variable that does not vary within a cluster but between, e.g. teacher experience.\\
\addlinespace
Hierarchical model & Model accounting for dependant observations relying on certain parameters ( within cluster) which in turn depend on other parameters (between cluster)\\
Fixed effect & Effects that are constant across all sample units, e.g. something that researchers control for and can repeat, such as  a teaching strategy (tutoring after class)\\
Random effect & Effects that are a source of random variation in the data, and whose levels are not fully sampled. e.g. test score tendency during academic year between students due to no controlled factors such as  genetic,family history\\
Mixed effect & Includes fixed and random effects, e.g. the fixed effect would be the treatment effect of a drug and the random effect would be the ID of the hospital where the patient is treated. Multilevel models typically accommodate for variability by including a separate group mean for each cluster e.g random intercept on hospitals. In addition to random intercepts, multilevel models can also include random coefficients and heterogeneous residual error variances across clusters (see e.g. @gelm06, @hox17 and @jong21).\\
ICC & The variability due to clustering is often measured by means of the intraclass coefficient (ICC). The ICC can be seen as the percentage
of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.\\
\addlinespace
Stratified intercept & \\
\bottomrule
\end{tabular}
\end{table}

In \proglang{R}, multilevel models may be fitted using package such as
\pkg{lme4} or \pkg{nlme}. For linear mixed-effects models, the function

@@

\hypertarget{concepts-in-missing-data}{%
\subsection{Concepts in missing data}\label{concepts-in-missing-data}}

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-2-1.pdf}

@@

Table 2: Concepts in missing data methods

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.1739}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.8261}}@{}}
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Concept}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Details}
\end{minipage} \\
\midrule()
\endhead
MCAR & Missing Completely At Random, where the probability to be missing
is equal \\
& across all data entries \\
MAR & Missing At Random, where the probability to be missing depends on
observed \\
& information \\
MNAR & Missing Not At Random (MNAR), where the probability to be
missing \\
& depends on unrecorded information, making the missingness
non-ignorable \\
& \citep{rubi76, meng94}. \\
& \\
\bottomrule()
\end{longtable}

\hypertarget{imputation-with-mice}{%
\subsection{Imputation with mice}\label{imputation-with-mice}}

@@

\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, leftrule=.75mm, toprule=.15mm, colframe=quarto-callout-color-frame, rightrule=.15mm, bottomrule=.15mm, colback=white, breakable, opacityback=0, left=2mm]

@@

\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, leftrule=.75mm, toprule=.15mm, colframe=quarto-callout-color-frame, rightrule=.15mm, bottomrule=.15mm, colback=white, breakable, opacityback=0, left=2mm]

TODO: @ Hanne i added a wide explanation about hte prediction matrix in
the Model Specification subsection, so you can check it and consider to
modify your box.. because i dont think is usefull to repeat information
across the paper\ldots{}

@@

Before diving into imputation, make sure your estimation model can
handle missing data. Sometimes, simpler methods like complete case
analysis might be suffice, especially if your missingness is low
(usually \textless5\%). There are scenarios in which specific Maximum
Likelihood (ML) estimation methods outperform Multiple Imputation (MI)
methods, for instance when the response variable is the sole incomplete
variable, mixed models demonstrate robustness to missing data under the
Missing at Random (MAR) assumption and with a correct
variance-covariance specification.\cite{molenberghs2007}.

@@

The imputation model can incorporate additional terms compared to the
main model without causing compatibility issues. For instance, it's
recommended to include the outcome variable in the imputation model for
prediction variables \cite{moons2006}. In cases where the outcome is
time-to-event, the Nelson-Aalen estimate of the time to the event should
be added as a covariate in the imputation model {[}REF{]}. Additionally,
including auxiliary variables, even if not part of the main model, can
be linked to the probability of missingness, improving the likelihood of
meeting the Missing at Random (MAR) assumption and enhancing estimation
efficiency \cite{hardt2012}.

@@

Recipes have been proposed for imputing incomplete level-1 and level-2
variables for hierarchical models \cite{buur18} (\(\S\) 7.10), which can
be convenient to follow when dealing with numerous variables
(interaction terms at different levels) and models with many random
effects that are prone to convergency problems due to overspecification
in the imputation model. These rules were designed to ensure
compatibility among the conditionally specified imputation models and
congeniality between the imputation and the main model. However, they
may not be applicable to all 2l imputation methods; for instance, the
2l.2stage imputation methods of the micemd package only allow the
inclusion of random predictor variables (2).

On the other hand, the formula option is useful in specifying complex
imputation models with polynomial terms or interactions and compared
with the prediction matrix method do not requires the inclusion of
additional terms as Just Another Variable (JAV) \cite{buur18}(\(\S\)
6.4).

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-7-1.pdf}

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-9-1.pdf}

@@

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-16-1.pdf}

@@

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-27-1.pdf}

@@

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-37-1.pdf}

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-39-1.pdf}

@@

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1605}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1111}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1235}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0988}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1358}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.2222}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0741}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0741}}@{}}
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Estimate
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
Std.Error
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
t.value
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
df
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
P(\textgreater\textbar t\textbar)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
RIV
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
FMI
\end{minipage} \\
\midrule()
\endhead
(Intercept) & -1.910 & 0.335 & -5.700 & 9617.768 & 0.000 & 0.021 &
0.021 \\
type & -0.347 & 0.179 & -1.941 & 256541.695 & 0.052 & 0.004 & 0.004 \\
age & 0.032 & 0.002 & 19.550 & 1965.628 & 0.000 & 0.047 & 0.046 \\
motor\_score2 & -0.576 & 0.070 & -8.195 & 14898.703 & 0.000 & 0.017 &
0.017 \\
motor\_score3 & -0.892 & 0.072 & -12.390 & 14760.294 & 0.000 & 0.017 &
0.017 \\
motor\_score4 & -1.272 & 0.074 & -17.203 & 10235.157 & 0.000 & 0.020 &
0.020 \\
pupil2 & 1.311 & 0.074 & 17.758 & 41.455 & 0.000 & 0.451 & 0.342 \\
pupil3 & 0.661 & 0.077 & 8.533 & 165.867 & 0.000 & 0.184 & 0.165 \\
ct2 & 0.783 & 0.075 & 10.461 & 264.558 & 0.000 & 0.140 & 0.130 \\
ct3 & 0.786 & 0.095 & 8.280 & 10.443 & 0.000 & 1.624 & 0.676 \\
\bottomrule()
\end{longtable}

@@

\begin{longtable}[]{@{}lr@{}}
\toprule()
& Estimate \\
\midrule()
\endhead
Intercept\textasciitilde\textasciitilde Intercept\textbar name &
0.083 \\
\bottomrule()
\end{longtable}

\hypertarget{obesity-data}{%
\subsection{Obesity data}\label{obesity-data}}

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-47-1.pdf}

@@

\begin{verbatim}
R> Nulmodel <- lme4::lmer(BMI ~ 1 + (1|Cluster), data = Obesity)
R> performance::icc(Nulmodel)
\end{verbatim}

\begin{verbatim}
# Intraclass Correlation Coefficient

    Adjusted ICC: 0.362
  Unadjusted ICC: 0.362
\end{verbatim}

@@

It suggests using `2l.2stage.bin' for the FAV variable and
`2l.2stage.norm' for the age variable. However, after inspecting the age
and weight density plot, we consider modifying its method to
`2l.2stage.pmm'. For the BMI variable, we employ deterministic
imputation.

\begin{verbatim}
meth_mar <-
  find.defaultMethod(
    Obesity,
    ind.clust = 1,
    I.small = 7,
    ni.small = 100,
    prop.small = 0.4
  )
meth_mar["BMI"] <- "~ I(Weight / (Height)^2)"
meth_mar["Age"] <- "2l.2stage.pmm"
meth_mar["Weight"] <- "2l.2stage.pmm"
\end{verbatim}

@@

\begin{verbatim}
pred_mar <- quickpred(Obesity)
pred_mar["Weight",] <- 2
pred_mar["Weight","Weight"] <- 0
pred_mar[, "Cluster"] <- -2 
pred_mar[, "Time"] <- 0
pred_mar[pred_mar == 1] <- 2

pred_mar[c("Height", "Weight"), "BMI"] <- 0
#plot_pred(pred_mar)
imp_mar <-
  mice(
    data = Obesity,
    method = meth_mar,
    predictorMatrix = pred_mar,
    printFlag = FALSE
  )
\end{verbatim}

\begin{verbatim}
summary(complete(imp_mar, "long")$Weight)
\end{verbatim}

\begin{verbatim}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  28.35   69.38   81.96   82.36   94.46  134.61 
\end{verbatim}

\begin{verbatim}
plot_trace(imp_mar, "Weight")
\end{verbatim}

@@

\begin{verbatim}
ggmice(imp_mar, aes(.imp, Weight)) + 
  geom_jitter()
\end{verbatim}

@@

@@

As before we use the `pmm', option but this time for the Heckman
imputation, this approach ensures that the imputed values remain within
the range of observable values. We then run the MNAR imputation model by
setting the pmm parameter to true.

\begin{verbatim}
imp_mnar <-
  mice(
    data = Obesity,
    method = meth_mnar,
    predictorMatrix = pred_mnar,
    pmm = T,
    printFlag = FALSE
  )
\end{verbatim}

@@

\begin{verbatim}
   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
  28.35   67.39   82.57   81.71   96.04  134.61 
\end{verbatim}

\begin{verbatim}
plot_trace(imp_mar, "Weight")
\end{verbatim}

@@

We proceed to compare the effects on the model. We run the analysis
model on each of the completed datasets as well as the dataset where the
incomplete values are removed (Complete Case analysis, CC).

We proceed to fit the mixed model by using the complete case database
and the imputed datasets (MAR,MNAR), and we compare the estimates of
each dataset by using a plot (the plot function can be found in the
appendix).

\begin{verbatim}
# Save all datasets in a list
list_data <- list(Obesity |> filter(complete.cases(Obesity)), # complete case
                  imp_mar, # imputed under mar
                  imp_mnar) # imputed under mnar

# Run mixed models on each dataset
list_models <- lapply(list_data, 
                   FUN = function(x) 
                     with(x,nlme::lme(BMI ~ Age + FamOb + Gender, 
                                      random =  ~ 1 + Age | Cluster)))
# Plot coefficients
plot_models(list_models,
             mod_name = c("Complete case", "MAR", "MNAR"))
\end{verbatim}

{\centering \includegraphics{manuscript_files/figure-pdf/models-1.pdf}
